---
output: 
  html_document: 
    keep_md: yes
---

**Reproducible Research Assignment 1**
=====================================

###**Loading and processing the data**

```{r}
activity <- read.csv("C:/Users/vgw52064/Desktop/Coursera/Reproducible Research/activity.csv")
str(activity)
summary(activity)
```

###**What is the mean of the total number of steps taken per day?**

First, total number of steps is calculated.

```{r}
totalsteps <- tapply(activity$steps, activity$date, FUN = sum, na.rm = TRUE)
summary(totalsteps)
```

Additionally, the mean and median values of the total steps taken per day is calculated
and formatted.
Lastly, a histogram of the total steps taken per day is generated.

```{r}
totalstepsMean <- mean(totalsteps)
Mean <- format(totalstepsMean, big.mark = ",", nsmall = 2)
totalstepsMed <- median(totalsteps)
Median <- format(totalstepsMed, big.mark = ",", nsmall = 2)

hist(totalsteps, ylim= c(0,30), main = "Total Number of Steps per Day", 
          xlab = "Steps", col = "purple")
     abline(v=totalstepsMed, col="pink", lwd=4)
     abline(v=totalstepsMean, col="lightblue", lwd=4)
     legend(x="topright", legend = c("Mean", "Median"), col = c("lightblue", "pink"), lwd = 4)

```


The mean of the total number of steps taken per day is **`r Mean`** and shown in blue in the graph.



The median of the toal number of steps taken per day is **`r Median`** and shown in pink in the graph.

###**What is the average daily activity pattern?**

First, the average number of steps taken per 5-minute intervals is determined.
A time series plot is generated and the maximum value is extracted.

```{r}
library(ggplot2)
avgint <- aggregate(x=list(steps = activity$steps), by =list(interval = activity$interval),
                    FUN = mean, na.rm=TRUE)

ggplot(data = avgint, aes(x=interval, y=steps)) + geom_line(color="purple") + xlab("5-minute Interval") + 
          xlim(c(0,2400)) + ylab("Average Number of Steps") + ggtitle("Average Daily Activity")

avgint[which.max(avgint$steps),]
```


The **835th** 5-minute interval contains the maximum average number of steps at a total of **206** steps. 


###**Missing Values**

The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
sum(is.na(activity))
```

Currently, there are **2,304** days/intervals where there are missing values (coded as NA) in the dataset. 

Missing values will be substituted by the mean value for that 5-minute interval and a new dataset with **no** missing values is created.

```{r}
average <- tapply(activity$steps, activity$interval, mean, na.rm=T)
filled <- transform(activity, steps=ifelse(is.na(steps), average, steps))
summary(filled)
sum(is.na(filled))
```

Total number of steps for the new dataset with no missing values is calculated.

```{r}
totalsteps.filled <- tapply(filled$steps, filled$date, FUN = sum)
summary(totalsteps.filled)
```

Additionally, the mean and median values of the total steps taken per day is calculated
and formatted for the new dataset.
Lastly, a histogram of the total steps taken per day is generated for the new dataset.

```{r}
totalstepsMean.filled <- mean(totalsteps.filled)
Mean.filled <- format(totalstepsMean.filled, big.mark = ",", nsmall = 2)
totalstepsMed.filled <- median(totalsteps.filled)
Median.filled <- format(totalstepsMed.filled, big.mark = ",", nsmall = 2)

hist(totalsteps.filled, ylim= c(0,30), main = "Total Number of Steps per Day", xlab = "Steps", col = "purple")
     abline(v=totalstepsMed.filled, col="yellow", lwd=4)
     abline(v=totalstepsMean.filled, col="lightblue", lwd=4, lty=2)
     legend(x="topright", legend = c("Mean", "Median"), col = c("lightblue", "yellow"), lwd = 4)

```


The mean of the total number of steps taken per day is **`r Mean.filled`** and shown in blue in the graph.



The median of the toal number of steps taken per day is **`r Median.filled`** and shown in pink in the graph.


Now that no NA values are present in the data, the mean and median are equal.

###**Differences in activity patterns between weekdays and weekends**

New factor variables are created "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
This data is then appended onto the dataset containing no NA values.

```{r}
days <- function(date){
     day <- weekdays(date)
     if(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
          return("Weekday") else if (day %in% c("Saturday", "Sunday"))
          return("Weekend") else stop("invalid date")
}
filled$date <- as.Date(filled$date)
filled$day <- sapply(filled$date, FUN= days)
```

Panel plot on the average number of steps taken on weekdays and weekends is shown below:

```{r}
library(ggplot2)
dayavg <- aggregate(steps~interval + day, data = filled, mean)

ggplot(data = dayavg, aes(x=interval, y=steps)) + geom_line(color="purple") + facet_grid(day ~.) +
          xlab("5-minute Interval") + xlim(c(0,2400)) + ylab("Average Number of Steps") + 
          ggtitle("Average Activity by Weekday or Weekend")

dayavg[which.max(dayavg$steps),]
```

The maximum average number of steps still occurs at the **835th** 5-minute interval, 
it is classified as a **weekday** and at a total of **230** steps. 
